Landau quantization of nearly degenerate bands

### 2018-09-19 paper.tex
Updates from ShareLaTeX

For simplicity, we assume that the Pauli matrices in \q{eq:Rashba-Hamiltonian} correspond to spin: $s_i{=}\hbar \sigma_i/2$. Let us evaluate the effective Hamiltonian $\calh(\bk)$  along the cyclotron trajectory $\frako_0$ with radius $k$. In the  energy basis of the Rashba Hamiltonian $H_R$ [cf.\ \q{eq:Rashba-Hamiltonian}], 
\e{\calh(\bk)=-\alpha k\tau_z{+}\f{g_s}{2}\mu_{\sma{B}}B\tau_x \red{-} \f{eB\hbar}{2m} (\tau_x-I) \epsilon_{\alpha\beta} k_{\red{\alpha}}\nabla_{\red{\beta}}\theta,\la{rashbaeffham}}
where $\tau_i$ are a new set of Pauli matrices with $\tau_z{=}{+} 1$ and ${-}1$ corresponding to the outer and inner orbits, respectively. $e^{i\theta_{\bk}}{=}(ik_x{-}k_y)/k$ is the angle of the in-plane spin-orbit field $({-}\alpha k_y,\alpha k_x,0)$, which winds once over $\frako_0$; the spin of our basis vectors likewise winds, in parallel (or anti-parallel) alignment with this field (see Fig. (\ref{fig:orbits})(b)). Terms proportional to the rate of spin rotation  ($\nabk \theta$) originate from the  non-abelian Berry connection. The first term of  $\calh$ corresponds to the spin-orbit-induced splitting of the energy-degenerate bands;
the remaining terms may be interpreted as B-induced deformations of the bands.   It is well-known that the Zeeman term tends to polarize spin in the direction ($\vec{z}$) of $\bB$; remarkably, the non-abelian Berry term   also has a polarizing effect, but with opposite sign. These two effects completely cancel for free electrons ($m{=}m_0$, with $m_0$ being free electron mass.), but not generically in band systems. The orbital moment term $B M_z$, which generally couples the two-band subspace (of interest) to all other bands, is absent in any simplified two-band model.
%Observe that  that $\calh$ is actually time-independent, which originates from a  that is unbroken by the field (aligned with the rotational axis). Consequently,

@@

\begin{figure}
\includegraphics[width=1.0\columnwidth]{RD.png}
% alpha=0.05
% beta=0.02
% m=0.01
\caption{Rashba-Dresselhaus model. (a) Typical cyclotron orbits. (b) Comparison of Landau levels calculated by Eq. (\ref{eq:rule}) (black lines) and numerical diagonalization of the Peierls-Onsager Hamiltonian  $H_{RD}(\bK)$ (red dots). Parameters for (b): $\alpha=0.05$(eV$\cdot$\AA), $m=0.01$($\hbar^2$/eV$\cdot$\AA$^2$), $\beta=0.02~(0.04)$(eV$\cdot$\AA) for left (right) panels. $B$ is measured from 2.9 and 1.0T respectively and is in unit $10^{-2}$T. (c) For various spin-orbit parameters $(\alpha,\beta)$ of the Rashba-Dresselhaus model and various field strength (satisfying $l^2S(E)\gg 1$), we plot  $\text{error}:=|\epsilon_\text{rule}-\epsilon_\text{exact}|/\var_c$ for a single Landau level that lies closest to energy $E=0.4eV$, where $\epsilon_\text{rule}$ is determined by Eq. (\ref{eq:rule}) and $\epsilon_{\text{exact}}$ is obtained by numerical diagonalization of the Peierls-Onsager Hamiltonian  $H_{RD}(\bK)$. We also plot in (d) $l^2\delta S^2/S$, which has a strong positive relation with error in (c). Parameters for (c-d): $m=0.01$($\hbar^2$/eV$\cdot$\AA$^2$). Units: $\alpha$($10^{-2}$eV$\cdot$\AA), $\beta$($10^{-2}$eV$\cdot$\AA), $Z$($10^{-3}$eV), $l^2 S$($10^2$).\label{fig:RD}}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1.0\columnwidth]{RZ.png}
\caption{(a) Trefoille-shaped cyclotron orbit for the Rashba-Zeeman model at the energy of the Dirac point. The orbit is drawn over the  {spin-orbit field} ($d_x,d_y$) of the Hamiltonian $H_{RZ}{=}d_0(\bk)I{+}d_x(\bk)\sigma_x{+}d_y(\bk)\sigma_y$, with $d_j$ obtainable from \q{eq:RZ-Hamiltonian}. The wavefunction along the inner (resp.\ outer) component of the trefoille orbit has its pseudospin aligned parallel (resp.\ anti-parallel) to the spin-orbit field. (b) Black dashed lines illustrate a zeroth-order approximation of the Landau levels (assuming Klein tunneling); the leading-order correction is indicated by red lines. The Dirac point lies at $10$meV. Parameters for (b): $m$: 0.02$\hbar^2$/eV$\cdot$\AA$^2$, $\alpha$: 0.05eV$\cdot$\AA, $Z_\parallel$: 0.001eV. (c)\textbf{ Can we add axes E vs kx vs ky for main figure, and another set of axes kx vs ky for inset?} A sketch of cyclotron orbits of Rashba Hamiltonian with in-plane magnetic field. (d) For various configurations ($\alpha$, $Z_\parallel$) of $H_{RZ}$, we plot in left panel $\text{error}:=|\epsilon_\text{rule}-\epsilon_\text{exact}|/\var_c$ for a single Landau level that is closest to $\epsilon_0$ in Eq. (\ref{whereisdiracpoint}), where $\epsilon_\text{rule}$ is determined by Eq. (\ref{eq:rule}) and $\epsilon_{\text{exact}}$ is obtained by numerical diagonalization of the Peierls-Onsager Hamiltonian $H_{RZ}(\bK)$. Corresponding $l^2 \delta S^2/S$ are plotted in right panel. We only plotted in region where $0.4<\epsilon_0<1.0$(eV).\footnote{$l^2S\gg 1$ is assured by requiring $\epsilon_0>0.4$; the reason of choosing $\epsilon_0<1.0$ is Dirac point lying at high energy requires too much harmonics in numerical diagonalization. This requirement is equivalent to $\sqrt{0.8m}\alpha/\hbar<Z<\sqrt{2m}\alpha/\hbar$, corresponding to the boundaries of the plots in (d).} Parameters for (d): $m$: 0.01$\hbar^2$/eV$\cdot$\AA$^2$, $l=100$\AA. \textbf{please choose formatting such that footnote is at the back} \label{fig:RZ}}
\end{figure}

@@

In the absence of symmetry,  eigenvalue-degeneracies of the unitary propagator $\A$ [cf. \q{eq:prop}] are attainable by tuning three real parameters. Analogously, it is well known (as the non-crossing rule in quantum theory[von Neumann,wigner]) that eigenvalue-degeneracies of complex, finite-dimensional matrix Hamiltonians are also attainable by tuning three real parameters.
 To prove the non-crossing rule for two-by-two unitaries, we begin by decomposing any such matrix as $\A{=}e^{i\phi}{\cal \bar{A}}$, such that  ${\cal \bar{A}}{\in}\text{SU}(2)$ being equal to $\pm I$ is a necessary and sufficient condition for an eigenvalue-degeneracy of $\A$. In the canonical parametrization of $\text{SU}(2)$ over the three-sphere $S^3$: 
\e{{\cal \bar{A}}{=}\matrixtwo{R_1{+}iR_2}{R_3{+}iR_4}{{-}R_3{+}iR_4}{R_1{-}iR_2},\;\; \sum_{j=1}^4R_j^2{=}1,\;\; R_j\in \mathbb{R},  \la{s3}}
${\cal \bar{A}}{=}{\pm}I$ lie on distinct points of $S^3$, and are attainable by tuning three real parameters: $R_3{=}R_4{=}0$, and the phase of $R_1{+}iR_2$ must equal $0$ or $\pi$. The conclusion that three parameters need be tuned remains valid for any  non-singular\footnote{By `non-singular parametrization', we mean a set of parameters related to those in \q{s3} by a non-singular coordinate transformation.} parametrization of SU$(2)$.

%If special unitaries were uniformly distributed over $S^3$, then encountering a spectral degeneracy for a randomly-selected special unitary would be an exceedingly rare event -- three parameters need be tuned to find either pole on $S^3$,  i.e., the codimension of degeneracy manifold is three. If only one  or two  parameters (in $\{R_1,\ldots,R_4\}$)  can be tuned, then in general we should not expect a degeneracy.

 To generalize the above discussion, let us parametrize ${\cal \bar{A}}$ by $d$ independent real numbers  $\{R_1,\ldots,R_{d}\}{:}{=}\bR $;  ${\cal \bar{A}}(\bR){=}{\pm}I$ defines a set of manifolds in $\bR$-space which we refer to as degeneracy manifolds; in the above example, these manifolds comprise two points on $S^3$. The codimension of the degeneracy manifold  is defined as $d$ minus the dimension of said manifold; alternatively stated, the codimension is the number of linearly-independent parameters needed to fine-tune a degeneracy. Here and henceforth, we adopt the shorthand that  \textit{parameters} will always refer to real parameters, and \textit{codimension} always refers to a codimension of a degeneracy manifold.

@@

It is well-known that symmetry may reduce codimension.[wigner,von neumann] The conceptually-simplest example is the approximate symmetry of band conservation, which exists in the adiabatic regime [cf.\ \s{sec:qtznrules}]. Recall from \q{phaseindependentorbit} that  $\cala$ becomes diagonal ($R_3{=}R_4{=}0$) in the basis of energy bands,
with phase[$R_1{\pm}iR_2]{=}{\pm} l^2\delta S/2$ plus field-independent corrections. Since this  phase alone  can be  tuned to attain $\cala{=}{\pm}I$, the codimension is unity.
As a function of $l^2$, we deduce that Landau-level quasi-crossings occur  with a periodicity of $2\pi/\delta S$. This periodicity  has a dual interpretation  in real space -- as a periodic increment of a single flux quantum for  the differential magnetic flux (between the two split orbits).

@@

\subsection{Unit codimension for the Rashba 2DEG}\la{sec:singleparameterrashba}

To exemplify the role of rotational symmetry in reducing codimension,
we return to the Rashba model [cf.\ \q{eq:Rashba-Hamiltonian}] absent Dresselhaus spin-orbit coupling and in-plane Zeeman field.  Owing to a continuous rotational symmetry (denoted $\mathfrak{c}_{\infty}$), $\calh(\bk)$ is a constant along the circular zeroth-order orbit [cf.\ \q{rashbaeffham}], and therefore the propagator simplifies to $\A{=}$exp$[{-}i\calh T_c/\hbar]$. The latter is expressible in the exponential form:
\e{\A =e^{-i2\pi(r_1\tau_x{+}r_2\tau_z)}, \;\;(r_1,r_2){=}\bigg(\f{g_{\sma{\text{eff}}}}{4}-\f1{2},\f1{2}-\f{\alpha k_{{E}}}{\var_c}\bigg).\notag} 
All degeneracies occur along circles of radius $\sqrt{r_1^2+r_2^2}{=}{j}/2$, with ${j}$ a nonnegative even (resp.\ positive odd) integer for  $\A{=}I$ (resp.\ $\A{=}{-}I$). In this exactly-soluble model [cf.\ \q{eq:Rashba-exact}], all  degeneracies (${j} {\ge} 0$) of $\A$ correspond to exact Landau-level degeneracies, rather than the near degeneracies described by \qq{llquasideg}{llquasidegB}.

That the ${j}{=}0$ manifold has codimension two can be understood from an elementary result in quantum mechanics: taking $\alpha{=}0$ \textit{and} $m{=}m_0$, we recover the Zeeman-split Landau levels of a free electron gas without spin-orbit coupling; all levels (except the lowest) are spin-degenerate owing to the equality of the Zeeman and cyclotron energies.[LandauLifshitzstatphysI]   All other manifolds ($j{>}0$) have unit codimension, and can be tuned by varying $B$ alone (with fixed $g_{\sma{\text{eff}}}$); these manifolds are analogous to line nodes (of Hamiltonian degeneracies) in 2D $\bk$-space -- a topic of recent interest.[citations]

\begin{figure}
\includegraphics[width=1.0\columnwidth]{dgn.png}
\caption{Degenerate manifold of Rashba-Dresselhaus model. (a) degeneracy manifold of $\A=-I$; (b) degeneracy manifold of $\A=I$. Units: $\alpha l^2$($400$eV$\cdot$\AA$^3$), $\beta$($400$eV$\cdot$\AA$^3$), $g_s'$(10). Parameters are $m=0.01\hbar^2$/(eV$\cdot$\AA$^2$) and $E=0.4$eV. \red{[For 3D pictures, I always put units in caption since there's no room for them in the pictures.]}\label{fig:dgn}}
\end{figure}

The Rashba model suffers from an oversimplification -- continuous rotational symmetry is not a symmetry of solids. That is to say, it does not belong in any crystallographic space group. It is possible for $\mathfrak{c}_{\infty}$-symmetric $\bk{\cdot}\bp$ models to approximate Bloch Hamiltonians (defined over the Brillouin torus) with a discrete, order-$N$ rotational symmetry (denoted $\mathfrak{c}_N$). To what extent are degeneracies of $\A$ (for $\mathfrak{c}_{\infty}$-symmetric systems) robust when $\mathfrak{c}_{\infty}$ is perturbatively reduced to $\mathfrak{c}_N$? For each of $N{=}2,3,4$ and $6$ (the allowed values in space groups), we find that $N{-}1$ out of every $N$ degeneracies (of $\A$) are robust, as proven in App. \ref{app:codimension}.

For illustration, let us reduce $\mathfrak{c}_{\infty}$ symmetry to  $\mathfrak{c}_2$  by adding the Dresselhaus spin-orbit term ($\propto \beta$) to the Rashba model [cf.\ \q{hamRD}].  Since $\A$ depends on $\alpha,\beta,l$ only through $\alpha l^2$ and $\beta l^2$, we consider the three-dimensional parameter space $(\alpha l^2,\beta l^2,g_{\sma{\text{eff}}})\in \mathbb{R}^3$. If a  line node (in the $\beta{=}0$ plane) were perturbatively stable to finite $\beta$, then it would be embedded in a two-dimensional degeneracy hypersurface of $\mathbb{R}^3$; if the planar line node were unstable, this embedding does not happen, i.e., the planar line node is simply a line node in $\mathbb{R}^3$. Our numerical study shows that half the planar line nodes (corresponding to $\A{=}I$) are stable, while the other half ($\A{=}{-}I$) destabilizes. In $\mathbb{R}^3$, we then have pairs of line nodes and hypersurfaces enclosed by increasingly larger pairs of line nodes and hypersurfaces, as illustrated in Fig. \ref{fig:dgn}.

To understand this pattern of stabilities, consider  the effective Hamiltonian $\calh$ in momentum independent basis (rather than spin-split basis used before)
\e{\calh=\alpha (k_{x}\sigma_{y}{-}k_{y}\sigma_{x})+\beta (k_{x}\sigma_{x}{-}k_{y}\sigma_{y})-\tf{g_s}{2}\mu_{B}B\sigma_z;}
with the Pauli matrices corresponding to spin operators. The two-fold rotational symmetry 
\e{\calh(\bk)=\sigma_z\calh(-\bk)\sigma_z, \;\: \bk(t)=-\bk\big(t+\tf{T_c}{2}\big)\in \frako_0,}
implies that the propagator over the time interval $[T_c/2,0]$ is symmetry-related to that over $[T_c,T_c/2]$:
\e{\A=\A_{T_c\leftarrow T_c/2}\A_{T_c/2\leftarrow 0}=\sigma_z{\A}_{T_c/2\leftarrow 0}\sigma_z {\A}_{T_c/2\leftarrow 0}.\la{eq:sigmazconstraint}}
The tracelessness of $\calh$ (at each $\bk$) implies that  ${\A}_{\sma{T_c/2\leftarrow 0}}$  has unit determinant. 
Utilizing the canonical parametrization for this $\text{SU}(2)$ matrix [Eq. (\ref{s3}) with ${\cal \bar{A}}$ replaced by ${\A}_{\sma{T_c/2\leftarrow 0}}$], we derive 
\e{
{\A}={\cal \bar{A}}=\matrixtwo{1{-}2R_2(R_2{-}iR_1)}{2iR_2(R_3{+}iR_4)}{{-}2iR_2(R_3{-}iR_4)}{1{-}2R_2(R_2{+}iR_1)} \label{proprashba}
}
with $\sum_{j=1}^4R_j^2{=}1$. We deduce from \q{proprashba} that $R_2{=}0$ is a necessary and sufficient condition for ${\cal \bar{A}}{=}I$. The corresponding degeneracy manifold is therefore a  unit-codimensional hypersurface which includes the free-electron limit: $\bR{=}(1,0,0,0)$. \red{[Discussion: I really hate introducing new symbols.]} \fig{fig:dgn}(b) illustrates the same hypersurface plotted with a different choice of coordinates: $(\alpha l^2,\beta l^2,g_s')$. On the other hand, ${\cal \bar{A}}{=}{-}I$ occurs if and only if $\bR{=}(0,1,0,0)$ -- a point in $\bR$-space. If $C_2$ were the only symmetry, then three parameters would be needed to tune ${\cal \bar{A}}{=}{-}I$. Generally, additional symmetry further constrains $\bR$, which may effectively reduce the number of tuning parameters.  This explains why the ${\cal \bar{A}}{=}{-}I$ manifolds, as plotted in Fig. \fig{fig:dgn}(a)], are line nodes instead of point nodes: these line nodes lie in either the $\alpha{=}0$ or the $\beta{=}0$ plane, both of which are $\mathfrak{c}_{\infty}$-symmetric.\footnote{To complete the explanation, \textbf{speculative:} the space of $C_{\infty}$-symmetric unitaries lie on lines in $\bR$-space that intersect $\bR{=}(0,1,0,0)$ [the point node]. A singular coordinate transformation to $\br$-space maps these $C_{inf}$-symmetric lines to $\mathfrak{c}_{inf}$-symmetric planes, and the point node to a line node.}

>>>>

For simplicity, we assume that the Pauli matrices in \q{eq:Rashba-Hamiltonian} correspond to spin: $s_i{=}\hbar \sigma_i/2$. Let us evaluate the effective Hamiltonian $\calh(\bk)$  along the cyclotron trajectory $\frako_0$ with radius $k$. In the  energy basis of the Rashba Hamiltonian $H_R$ [cf.\ \q{eq:Rashba-Hamiltonian}], 
\e{\calh(\bk)=-\alpha k\tau_z{+}\f{g_s}{2}\mu_{\sma{B}}B\tau_x {-} \f{eB\hbar}{2m} (\tau_x-I) \epsilon_{\alpha\beta} k_{\alpha}\nabla_{\beta}\theta,\la{rashbaeffham}}
where $\tau_i$ are a new set of Pauli matrices with $\tau_z{=}{+} 1$ and ${-}1$ corresponding to the outer and inner orbits, respectively. $e^{i\theta_{\bk}}{=}(ik_x{-}k_y)/k$ is the angle of the in-plane spin-orbit field $({-}\alpha k_y,\alpha k_x,0)$, which winds once over $\frako_0$; the spin of our basis vectors likewise winds, in parallel (or anti-parallel) alignment with this field (see Fig. (\ref{fig:orbits})(b)). Terms proportional to the rate of spin rotation  ($\nabk \theta$) originate from the  non-abelian Berry connection. The first term of  $\calh$ corresponds to the spin-orbit-induced splitting of the energy-degenerate bands;
the remaining terms may be interpreted as B-induced deformations of the bands.   It is well-known that the Zeeman term tends to polarize spin in the direction ($\vec{z}$) of $\bB$; remarkably, the non-abelian Berry term   also has a polarizing effect, but with opposite sign. These two effects completely cancel for free electrons ($m{=}m_0$, with $m_0$ being free electron mass.), but not generically in band systems. The orbital moment term $B M_z$, which generally couples the two-band subspace (of interest) to all other bands, is absent in any simplified two-band model.
%Observe that  that $\calh$ is actually time-independent, which originates from a  that is unbroken by the field (aligned with the rotational axis). Consequently,

@@

\begin{figure}
\includegraphics[width=1.0\columnwidth]{RD.png}
% alpha=0.05
% beta=0.02
% m=0.01
\caption{Rashba-Dresselhaus model. (a) Typical cyclotron orbits. (b) Comparison of Landau levels calculated by Eq. (\ref{eq:rule}) (black lines) and numerical diagonalization of the Peierls-Onsager Hamiltonian  $H_{RD}(\bK)$ (red dots). Parameters for (b): $\alpha=0.05$(eV$\cdot$\AA), $m=0.01$($\hbar^2$/eV$\cdot$\AA$^2$), $\beta=0.02~(0.04)$(eV$\cdot$\AA) for left (right) panels. (c) For various spin-orbit parameters $(\alpha,\beta)$ of the Rashba-Dresselhaus model and various field strength (satisfying $l^2S(E)\gg 1$), we plot  $\text{error}:=|\epsilon_\text{rule}-\epsilon_\text{exact}|/\var_c$ for a single Landau level that lies closest to energy $E=0.4eV$, where $\epsilon_\text{rule}$ is determined by Eq. (\ref{eq:rule}) and $\epsilon_{\text{exact}}$ is obtained by numerical diagonalization of the Peierls-Onsager Hamiltonian  $H_{RD}(\bK)$. We also plot in (d) $l^2\delta S^2/S$, which has a strong positive relation with error in (c). Parameters for (c-d): $m=0.01$($\hbar^2$/eV$\cdot$\AA$^2$). Units: $\alpha$($10^{-2}$eV$\cdot$\AA), $\beta$($10^{-2}$eV$\cdot$\AA), $Z$($10^{-3}$eV), $l^2 S$($10^2$).\label{fig:RD}}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1.0\columnwidth]{RZ.png}
\caption{(a) Trefoille-shaped cyclotron orbit for the Rashba-Zeeman model at the energy of the Dirac point. The orbit is drawn over the  {spin-orbit field} ($d_x,d_y$) of the Hamiltonian $H_{RZ}{=}d_0(\bk)I{+}d_x(\bk)\sigma_x{+}d_y(\bk)\sigma_y$, with $d_j$ obtainable from \q{eq:RZ-Hamiltonian}. The wavefunction along the inner (resp.\ outer) component of the trefoille orbit has its pseudospin aligned parallel (resp.\ anti-parallel) to the spin-orbit field. (b) Black dashed lines illustrate a zeroth-order approximation of the Landau levels (assuming Klein tunneling); the leading-order correction is indicated by red lines. The Dirac point lies at $10$meV. Parameters for (b): $m$: 0.02$\hbar^2$/eV$\cdot$\AA$^2$, $\alpha$: 0.05eV$\cdot$\AA, $Z_\parallel$: 0.001eV. (c) A sketch of cyclotron orbits of Rashba Hamiltonian with in-plane magnetic field. (d) For various configurations ($\alpha$, $Z_\parallel$) of $H_{RZ}$, we plot in left panel $\text{error}:=|\epsilon_\text{rule}-\epsilon_\text{exact}|/\var_c$ for a single Landau level that is closest to $\epsilon_0$ in Eq. (\ref{whereisdiracpoint}), where $\epsilon_\text{rule}$ is determined by Eq. (\ref{eq:rule}) and $\epsilon_{\text{exact}}$ is obtained by numerical diagonalization of the Peierls-Onsager Hamiltonian $H_{RZ}(\bK)$. Corresponding $l^2 \delta S^2/S$ are plotted in right panel. We only plotted in region where $0.4<\epsilon_0<1.0$(eV).\footnote{$l^2S\gg 1$ is assured by requiring $\epsilon_0>0.4$; the reason of choosing $\epsilon_0<1.0$ is Dirac point lying at high energy requires too much harmonics in numerical diagonalization. This requirement is equivalent to $\sqrt{0.8m}\alpha/\hbar<Z<\sqrt{2m}\alpha/\hbar$, corresponding to the boundaries of the plots in (d).} Parameters for (d): $m$: 0.01$\hbar^2$/eV$\cdot$\AA$^2$, $l=100$\AA. \textbf{please choose formatting such that footnote is at the back} \label{fig:RZ}}
\end{figure}

@@

In the absence of symmetry,  eigenvalue-degeneracies of the unitary propagator $\A$ [cf. \q{eq:prop}] are attainable by tuning three real parameters. Analogously, it is well known (as the non-crossing rule in quantum theory[von Neumann,wigner]) that eigenvalue-degeneracies of complex, finite-dimensional matrix Hamiltonians are also attainable by tuning three real parameters.
 To prove the non-crossing rule for two-by-two unitaries, we begin by decomposing any such matrix as $\A{=}e^{i\phi}{\cal \bar{A}}$, such that  ${\cal \bar{A}}{\in}\text{SU}(2)$ being equal to $\pm I$ is a necessary and sufficient condition for an eigenvalue-degeneracy of $\A$. In the canonical parametrization of $\text{SU}(2)$ over the three-sphere $S^3$: 
\e{{\cal \bar{A}}{=}\matrixtwo{r_1{+}ir_2}{r_3{+}ir_4}{{-}r_3{+}ir_4}{r_1{-}ir_2},\;\; \sum_{j=1}^4r_j^2{=}1,\;\; r_j\in \mathbb{R},  \la{s3}}
${\cal \bar{A}}{=}{\pm}I$ lie on distinct points of $S^3$, and are attainable by tuning three real parameters: $r_3{=}r_4{=}0$, and the phase of $r_1{+}ir_2$ must equal $0$ or $\pi$. The conclusion that three parameters need be tuned remains valid for any  non-singular parametrization of SU$(2)$.

%If special unitaries were uniformly distributed over $S^3$, then encountering a spectral degeneracy for a randomly-selected special unitary would be an exceedingly rare event -- three parameters need be tuned to find either pole on $S^3$,  i.e., the codimension of degeneracy manifold is three. If only one  or two  parameters (in $\{r_1,\ldots,r_4\}$)  can be tuned, then in general we should not expect a degeneracy.

 To generalize the above discussion, let us parametrize ${\cal \bar{A}}$ by $d$ independent real numbers  $\{r_1,\ldots,r_{d}\}{:}{=}\br $;  ${\cal \bar{A}}(\br){=}{\pm}I$ defines a set of submanifolds in $\br$-space which we refer to as degeneracy manifolds; in the above example, these manifolds comprise two points on $S^3$. The codimension of the degeneracy manifold  is defined as $d$ minus the dimension of said manifold; alternatively stated, the codimension is the number of linearly-independent parameters needed to fine-tune a degeneracy. Here and henceforth, we adopt the shorthand that  \textit{parameters} will always refer to real parameters, and \textit{codimension} always refers to a codimension of a degeneracy manifold.

@@

It is well-known that symmetry may reduce the codimension of Hamiltonians, e.g., an antiunitary symmetry may restrict complex Hamiltonians to real Hamiltonians, thus reducing codimension from three to two.[wigner,von neumann] In our context, an analog would be the symmetry of band conservation in the adiabatic regime [cf.\ \s{sec:qtznrules}], which restricts $\cala$ to have diagonal form  ($r_3{=}r_4{=}0$) in the basis of energy bands [cf.\ \q{phaseindependentorbit}]. The phase of $r_1{\pm}ir_2$ equals ${\pm} l^2\delta S/2$ plus field-independent corrections. Since this  phase alone  can be  tuned to attain $\cala{=}{\pm}I$, the codimension is unity. As a function of $l^2$, we deduce that Landau-level quasi-crossings occur  with a periodicity of $2\pi/\delta S$. This periodicity  has a dual interpretation  in real space -- as a periodic increment of a single flux quantum for  the differential magnetic flux (between the two split orbits).

@@

\subsection{Codimension reduction for the Rashba 2DEG}\la{sec:singleparameterrashba}

In this and the subsequent section [\s{sec:rotsymmbreakdown}], we will demonstrate how rotational symmetry reduces codimension in the Landau-level problem. Before delving into this, it is instructive to consider the conceptually-simplest analog in differential geometry: if we parametrize the 2D plane by Cartesian coordinates $(x,y){\in}\R^2$, it is apparent that two parameters must be tuned to arrive at the origin. If however the origin is the center of a rotation symmetry, we may instead utilize polar coordinates and tune a single parameter (the radial coordinate) to arrive at $x{=}y{=}0$. We see that rotational symmetry reduces the codimension of the origin from two to one. Here, and in the  Landau-level problem, we define codimension with respect to a \textit{symmetrically-chosen coordinates} (if symmetry exists). We do not require that the symmetric coordinates is globally defined, e.g., polar coordinates is singular at $x{=}y{=}0$.

\subsubsection{With continuous rotational symmetry}\la{sec:ctsrot}

%To exemplify the role of rotational symmetry in reducing codimension, we return to the Rashba model [cf.\ \q{eq:Rashba-Hamiltonian}] absent Dresselhaus spin-orbit coupling and in-plane Zeeman field.  Owing to a continuous rotational symmetry (denoted $\mathfrak{c}_{\infty}$), $\calh(\bk)$ is a constant along the circular zeroth-order orbit [cf.\ \q{rashbaeffham}], and therefore the propagator simplifies to $\A{=}$exp$[{-}i\calh T_c/\hbar]$. The latter is expressible in the exponential form:
%\e{\A =e^{-i2\pi(r_1\tau_1{+}r_3\tau_3)}, \;\;(r_1,r_3){=}\bigg(\f{g_{\sma{\text{eff}}}}{4}-\f1{2},\f1{2}-\f{\alpha k_{{E}}}{\var_c}\bigg).\notag} 
%All degeneracies occur along circles of radius $\sqrt{r_1^2+r_3^2}{=}{j}/2$, with ${j}$ a nonnegative, even (resp.\ positive, odd) integer for  $\A{=}I$ (resp.\ $\A{=}{-}I$). In this exactly-soluble model [cf.\ \q{eq:Rashba-exact}], all  degeneracies (${j} {\ge} 0$) of $\A$ correspond to exact Landau-level degeneracies, rather than the near degeneracies described by \qq{llquasideg}{llquasidegB}.

%That the ${j}{=}0$ manifold has codimension two can be understood from an elementary result in quantum mechanics: taking $\alpha{=}0$ \textit{and} $m{=}m_0$, we recover the Zeeman-split Landau levels of a free electron gas without spin-orbit coupling; all levels (except the lowest) are spin-degenerate owing to the equality of the Zeeman and cyclotron energies.[LandauLifshitzstatphysI]   All other manifolds ($j{>}0$) have unit codimension, and can be reached by varying $B$ alone (with fixed $g_{\sma{\text{eff}}}$).

To understand why rotational symmetry reduces codimension in the Landau-level problem, we return to the Rashba model [cf. \ \q{eq:Rashba-Hamiltonian}] absent Dresselhaus spin-orbit coupling and in-plane Zeeman field. This model has a continuous rotational symmetry (denoted $\mathfrak{c}_{\infty}$), which manifests as the constancy of $\calh(\bk){:}{=}\sum_{j=1}^3r_j\tau_j$ over the circular zeroth-order orbit [cf.\ \q{rashbaeffham}]; here, $\br{=}(r_1,r_2,r_3)$ is a real three-vector and $\{\tau_j\}$ a set of Pauli matrices.  Consequently, the propagator  simplifies to  the exponential form:
\e{\A{=}e^{-i2\pi\sum_{j}r_j\tau_j}, \;\;(r_1,r_3){=}\bigg(\f{g_{\sma{\ext{eff}}}}{4}{-}\f1{2},\f1{2}{-}\f{\alpha k_{\sma{E}}}{\var_c}\bigg).\la{expform}} 
$r_2{=}0$ in the Rashba model, but nonzero $r_2$ corresponds to a symmetry-allowed deformation that we allow for greater generality -- any $SU(2)$ matrix can be expressed in the exponential form [cf.\ \q{expform}]  for some (possibly non-unique) $\br$, which are the symmetric  coordinates.

%\textbf{[perhaps a footnote explaining this is a non-singular parametrization if no symmetry is imposed]}

All degeneracies of $\A$ occur on spheres in $\br$-space with radii equal to an integer ($j$)  multiple of $1/2$;  ${j}$ equals a nonnegative, even (resp.\ positive, odd) integer for  $\A{=}I$ (resp.\ $\A{=}{-}I$).  The zero-radius sphere (i.e., a point) has codimension three, as would be expected if symmetry were absent.    
This point degeneracy can be understood from an elementary result in quantum mechanics: taking $\alpha{=}r_2{=}0$ \textit{and} $m{=}m_0$, we recover the Zeeman-split Landau levels of a free electron gas without spin-orbit coupling; all levels (except the lowest) are spin-degenerate owing to the equality of the Zeeman and cyclotron energies.[LandauLifshitzstatphysI]

All other spheres are unit-codimensional surfaces (known as hypersurfaces), and can be reached by varying $B$ alone. Equivalently stated, at fixed energy there exists a series of Landau-level quasi-crossings as a function of $B$. The stability of hypersurfaces rely on  $\mathfrak{c}_{\infty}$ symmetry. Moreover, the hypersurfaces separate $\br$-space into infinitely-many, distinct connected components, i.e., the zeroth homotopy group $\pi_0$ comprises an infinite set. Each hypersurface may thus be viewed as a symmetry-protected, topological defect, which separates domains in $\br$-space where Landau levels are non-quasidegenerate.

This discussion is reminiscent of topological (semi)metals which host Dirac-Weyl points and/or line nodes (of Bloch-Hamiltonian degeneracies). Weyl points in 3D $\bk$-space have codimension three and do not require any crystallographic point-group symmetry for their stability. Just like Weyl points, the point degeneracy of $\A$ is associated with a nontrivial Chern number (defined as an integral of the Berry curvature\footnote{The curvature is defined with respect to a line bundle formed from the nondegenerate eigenvector of $\cala$.} over a Gaussian surface surrounding the point). On the other hand,  both Dirac points (in 2D $\bk$-space, e.g., graphene) and  line nodes (in 3D $\bk$-space) have codimension two, hence their existence relies on point-group symmetries.

\begin{figure}
\includegraphics[width=1.0\columnwidth]{dgn.png}
\caption{Degeneracy manifolds of the Rashba-Dresselhaus model, with units: $\alpha l^2$($400$eV$\cdot$\AA$^3$), $\beta$($400$eV$\cdot$\AA$^3$), $g_{\sma{\text{eff}}}$(10), and parameters: $m=0.01\hbar^2$/(eV$\cdot$\AA$^2$) and $E=0.4$eV. (a) illustrates the degeneracy manifold of $\A=-I$, which lie confined to  the $\alpha{=}0$ and $\beta{=}0$ planes. Each $\mathfrak{c}_{\infty}$-symmetric plane contains an infinite set of concentric, circular line degeneracies. Every circle (except for the smallest in either plane)  intersects two other circles in an orthogonal plane, such that the entire line node is path-connected. (b) The degeneracy manifold of $\A{=}I$ is a path-connected hypersurface.  \label{fig:dgn}}
\end{figure}

\subsubsection{With discrete rotational symmetry}\la{sec:disrot}

The Rashba model suffers from an oversimplification -- continuous rotational symmetry is not a symmetry in any crystallographic space group. It is possible for $\mathfrak{c}_{\infty}$-symmetric $\bk{\cdot}\bp$ models to approximate Bloch Hamiltonians (defined over the Brillouin torus) with a discrete, order-$N$ rotational symmetry (denoted $\mathfrak{c}_N$). How robust are the above-mentioned Landau-level quasi-crossings  (over the $B$ axis) when $\mathfrak{c}_{\infty}$ symmetry is perturbatively reduced to $\mathfrak{c}_N$? For any  integer $N{\geq}2$, we find that $N{-}1$ of every $N$ quasi-crossings  are perturbatively stable, as proven in App. \ref{app:codimension}.
